NumPy---Array Computation for Python

### 2019-12-24 body.tex
Use section numbers
Restructure based on feedback from Ralf
This is just a start.  We still need to cleanup
the prose.

\section*{Introduction}

@@

\subsection*{History}

@@

\subsection*{Funding}

@@

\subsection*{Current status}

% (maturity, users)

As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently and informally until 2017 when,
to facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
Since then there have been 19 proposed NEPS---6 have been implemented,
4 have been accepted and are being implemented, 4 are under
consideration, 3 have been deferred or superseded, and 2 have been rejected
or withdrawn.

The first activities organized for the new grant were a NEP development
sprint, a weekly community call, and a meeting to write a draft roadmap.

Since the roadmap was meant to represent community consensus on future
development, it was presented for comment at the annual SciPy
conference during a so-called ``Birds of a Feather'' session, attended
by more than 100 people.  One additional meeting was held to discuss
specific enhancements, such as the development of a new data-type
system.  The final roadmap was vetted by the community via online
discussion before being accepted.

Weekly community calls alternate between triage and
higher level discussion.  The calls not only involve developers from
the community, but provide a venue for vendors and other external
groups to provide input.  For example, after Intel produced a forked
version of NumPy, one of their developers joined a call to discuss
community concerns.

NumPy plays a central role in building and standardizing much of the scientific
Python community infrastructure.
The docstring standard mentioned in the history section is now widely adopted.
We are also now using the NEP system as a way to help coordinate the larger
scientific Python community.
% https://numpy.org/neps/nep-0029-deprecation_policy.html
For example, in NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem adopt a
common ``time window-based'' policy for support of Python and NumPy versions.
This standard will simplify downstream project and release planning.

\section*{Project organization and community}

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these, 17 maintainers were active in
2019, 4 of whom were paid to work on the project full-time.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through GitHub
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
with each release attracting more than a hundred new contributors.

@@

% -- The paragraphs above and below can be tightened up a bit still --

The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
PRs merged into the NumPy master branch.
In this plot PRs opened by current and old maintainers are indicated.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.
Over the course of its history, NumPy has attracted PRs by 823 contributors.
However, its development relies heavily on a relatively small number
of active maintainers, who share more than half of the contributions among
themselves.

% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
% https://github.com/numpy/numpy/pull/6352
To formalize its unwritten development policy,
NumPy adopted an official Governance Document on October~5,
2015\cite{NumPyProjectGovernance}.
Project decisions are usually made by consensus of interested contributors.
This means for many thing everyone is entrusted with veto power.
A Steering Council, currently composed of 12~members, facilitates this
process and oversees daily development of the project by contributing code
and reviewing contributions from the community.

% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
% https://github.com/numpy/numpy/pull/11865
NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
In brief, we strive to:
\emph{be open};
\emph{be empathetic, welcoming, friendly, and patient};
\emph{be collaborative};
\emph{be inquisitive}; and
\emph{be careful in the words that we choose}.
The Code of Conduct also specifies how breaches can be reported and outlines
the process for responding to such reports.

\section*{Library organization}

@@

\subsection*{Core}

@@

\subsection*{Computing libraries}

@@

\subsection*{Infrastructure libraries}

@@

\section*{Recent technical improvements}

The NumPy codebase has been actively developed for over two decades (first as
Numeric and NumArray).
Given the complexity of the codebase and the massive number of projects depending
on it, large changes require careful planning and substantial work.
Moreover, with the now prevalent need for distributed computing and the
emergence of specialized hardware such as GPUs, many new usecases have
arisen, which NumPy was not originally designed for.
With the recent infusion of funding and the newly formalized NEP process in
place we have been able to tackle a number of important large scale changes.
We highlight two of those below, as well as changes made to our testing
infrastructure to support hardware platforms used in large scale computing.

\subsection*{The Array Function Protocol}

@@

\subsection*{Random}

@@

\subsection*{Testing on multiple architectures}

@@

\section*{Discussion}

>>>>

\section{Introduction}

@@

\subsection{History}

@@

\subsection{Funding}

@@

\section{Organization and infrastructure}

@@

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these, 17 maintainers were active in
2019, 4 of whom were paid to work on the project full-time.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through GitHub
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
with each release attracting more than a hundred new contributors.
Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
master branch.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.

@@

\subsection{Core functionality}

@@

\subsection{Computing libraries}

@@

\subsection{Infrastructure libraries}

@@

\section{Recent technical improvements}

The NumPy codebase has been actively developed for over two decades (first as
Numeric and NumArray).
Given the complexity of the codebase and the massive number of projects depending
on it, large changes require careful planning and substantial work.
As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently and informally until 2017 when,
to facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
%Since then there have been 19 proposed NEPS---6 have been implemented,
%4 have been accepted and are being implemented, 4 are under
%consideration, 3 have been deferred or superseded, and 2 have been rejected
%or withdrawn.

The first activities organized for the new grant were a NEP development
sprint, a weekly community call, and a meeting to write a draft roadmap.
Since the roadmap was meant to represent community consensus on future
development, it was presented for comment at the annual SciPy
conference during a so-called ``Birds of a Feather'' session, attended
by more than 100 people.  One additional meeting was held to discuss
specific enhancements, such as the development of a new data-type
system.  The final roadmap was vetted by the community via online
discussion before being accepted.

%Weekly community calls alternate between triage and
%higher level discussion.  The calls not only involve developers from
%the community, but provide a venue for vendors and other external
%groups to provide input.  For example, after Intel produced a forked
%version of NumPy, one of their developers joined a call to discuss
%community concerns.

%NumPy plays a central role in building and standardizing much of the scientific
%Python community infrastructure.
%The docstring standard mentioned in the history section is now widely adopted.
%We are also now using the NEP system as a way to help coordinate the larger
%scientific Python community.
%% https://numpy.org/neps/nep-0029-deprecation_policy.html
%For example, in NEP 29, we recommend, along with leaders from various other
%projects, that all projects across the Scientific Python ecosystem adopt a
%common ``time window-based'' policy for support of Python and NumPy versions.
%This standard will simplify downstream project and release planning.

%Moreover, with the now prevalent need for distributed computing and the
%emergence of specialized hardware such as GPUs, many new usecases have
%arisen, which NumPy was not originally designed for.
With the recent infusion of funding and the newly formalized NEP process in
place we have been able to tackle a number of important large scale changes.
We highlight two of those below, as well as changes made to our testing
infrastructure to support hardware platforms used in large scale computing.

\subsection{The Array Function Protocol}

@@

\subsection{Random}

@@

\subsection{Testing on multiple architectures}

@@

\section{Discussion}
